### 2024-06-28 thesis/ms-fletcher-porter.tex
Moar work

In order to study mechatronics systems, it is necessary to design and develop physical research testbeds. This is especially the case in mechatronic systems with fluid components, as their high-fidelity modelling is challenging and computationally expensive. When doing research with hardware systems, before a single research question can be asked of the hardware, researchers must assemble mechanical components, hook up power electronics, and route hoses for the hydraulic and pneumatic systems. Then the most complex of them all, the control system, must be added. Its tangle of spindly wires reaches all throughout the machine, meeting at the computer, the single most complex component of any fluid power system. The requirements placed on it are immense. In high-power systems, safety invariants must be held up perfectly lest disaster strike, and close monitoring is necessary due to the low stiffness of fluid systems compared to mechanical ones.

@@

Dolores is a full-scale testbed in Aalto University's Fluid Power Laboratory, seen in figure~\ref{fig:dolores0} to study the application of electric hydrostatic actuation in an articulated-steering linkage for a front-loading mining vehicle~\cite{Hermansson2021, Helduser1999, Dudzinski1989}. Its control system commissioning was completed in 2021, but come 2023, an industry partner in the experiment requested hardware changes that demanded a substantial re-architecture of the control system. This case study will discuss the mechatronic tech stacks behind the old and the new systems.

@@

Also interesting in the architecture of a control system is what development environments can be constructed for it.  The development environment is distinct from the heretofore discussed mech stack since it is not a part of the function of the running system, but instead is means which an engineer uses to turn an idea into assets for a mechatronic system.  For a given tech stack, any development environment generally can in principle yield the same assets, but a good tech stack may lend itself to higher quality, more consistent output, faster development cycles, or at least to be enjoyable for individual engineers to use.

For the control layer, I wrote the code in the Sublime Text text editor~\cite{SublimeText} and configured it to automatically format the source code according to standards for the Rust language using the \verb|rustfmt| utility~\cite{rustfmt}.  Other text editors like Emacs, Vim, or VSCodium would have also been good choices, but Sublime is what is easiest for me.  This was done on a Windows workstation, but it was compiled and run on the Linux control computer.  To achieve this, I checked the changes on the workstation into a \verb|git| version control repository and pushed them to the control computer\footnote{A small issue that comes up is that \verb|git| generally doesn't allow users to push changes to a branch of a repository that the local user has checked out, which is the case here~\cite{GitConfigReceiveDenyCurrentBranch}.  This is to prevent data loss.  To enable this, \verb|git| can be configured to update the local branch only if there is no possibility of data loss by executing \verb|git config receive.denyCurrentBranch updateInstead|.}.  Then, though a remote terminal, I executed commands to compile and run the program.  This became quite annoying with how many steps there are, so I use the utility \verb|cargo-watch| to automatically execute the compile and run commands on the control computer when there are updates\footnote{The command was similar to \verb|cargo watch -x 'run -- eth0'|.}~\cite{CargoWatch}.  This still is somewhat tedious and so requires more development.  It also would have been mitigated by developing in some kind of Linux environment or by using a containerized environment like would be provided by Docker or Vagrant, which would have allowed running some tests on the development machine~\cite{DockerDev,VagrantDev}.  Most of the middleware was prepared with the Raspberry Pi OS image that was installed while comissioning the control computer, but there was some additional configuration, e.g.\ to install \verb|git|, that was done with the OS's \verb|apt| package management utility~\cite{apt}.  As well, library depndencies for the control program like Ethercrab were managed with the Rust \verb|cargo| utility~\cite{Cargo}.  The computer, once commissioned, did not require any development.  The IO layer's development was aided by a four-channel oscilliscope and a multimeter to validate electrical connections and ensure signal were propogating as expected.  It was also interesting to observe the timing jitter of the program, which was generally very small, on the order of microseconds.

The in-depth consideration of the development environment was possible because of the open nature of all the tools considered.  Every tool, both a part of the control system and the development environment, have interfaces that enable and encourage users to compose or develop tools to make an environment that meets their needs.  This same discussion could not be had with regards to Dolores since its PLC-based system was very closed off, and any interfacing demanded mastering of low-level communicaiton protocols.  The development environment was simply whatever Beckhoff blessed; no more, no less.

@@

The Solid Mechanics Laboratory at Aalto University conducts experiments to identify the properties of engineering materials under load particularly studying phenomena ranging from the micron to the meter scale.  As is typical for laboratories studying solid mechanics, they do this by subjecting test samples to a load schedule using a servo-controlled hydraulic cylinder.  They have several testbeds, each with an instance of a common control system.  This had been controlled by a workstation-commanded DAQ that communicated with a motor controller and a strain gauge amplifier, but new experiments and the desire for faster, better research cycles called for improvements to the control system.

The requirements that drive the design of this control system are mainly related to timing.  There are three compelling design parameters when it comes to timing.  The first is frequency or PID speed; how often in time do control and measurement samples occur?  According to Nyquist, the higher the frequency, the better the control performance~\cite{Nyquist1928}.  Latency, or how old a measurement sample is when it makes it to the PID loop, measured in samples or PID cycles, is also important to control performance.  Thoen shows that in some cases, a system with low sample frequency but low latency can outperform those with high sample frequency and high latency~\cite{Thoen2021}.  Jitter specifies how accurately the sample frequency is kept.  It doesn't impact control performance much, but for high-frequency systems, jitter can be important for compatibility to ensure synchrounous operation.  Its specification and control is generally the domain of analog circuit specialists instead of mechatronic generalists.

@@

The requirement for the IO system is that it be low latency.  Up to here the IO latency has been presented as given, but in reality the latency and PID
frequency specifications would be driven by the IO latency.  There are two separate IO systems, one for the strain gauges, which require specialized data acquisition to afford the quality of data needed for solid mechanics experiments.  Each instance of the control system has an HBK QuantumX amplifier, which introduces $\qty{0.3}{\milli\second}$ of latency~\cite{HBKQuantumX}.  For other IO, like signals to and from the servo motor driving the hydraulic pump or monitoring the state of the hydraulic circuit, there needed to be some analog IO ports.  The analog/digital and digital/analog converters contribute notable latency.  For the NI C-Series IO modules that were selected, analog output contributed $\approx\qty{6}{\micro\second}$ of latency per output channel and analog input added $\approx\qty{58}{\micro\second}$ of latency per two channels~\cite{NI-9218,NI-9266}.  This combined in some multiples and the IO, with conditioning, was measured to have the previously stated $ \qtyrange[range-units=single,range-phrase=..]{200}{400}{\micro\second}$ of latency.

The computer system Is interesting to discuss on account of the relatively high PID frequency requirement.  A frequency of $\qty{5}{\kilo\hertz}$ means that all computations for a loop must fit within $\qty{.2}{\milli\second}$.  Each CPU instruction for Intel microarchitectures usually takes ones or tens of clock cycles to execute, or even hundreds for heavy insructions~\cite{Abel19a}.  This means that on a computer with a $\qty{1.3}{\giga\hertz}$ clock like that which this thesis is being written with there can only be some thousands of instructions run per PID cycle.  This will include very fast arithmetic instructions like the additions or multiplications needed for the PID computation, but also commands to fetch data from the CPU's IO ports, which tend to be relatively slow since the signals need to be read and moved into a CPU register or memory location which have their own latency separate from the instruction latency.  On an embedded Linux or Windows system, all this would have to occur while sharing CPU time with other processes.  This isn't to say it's not possible to execute this with embedded Linux or Windows---Beckhoff and NI both provide off-the-shelf solutions for this---but performance issues can often distract from implementing business logic.  This can be mitigated with a bare metal system like was used for the above soft robots, but there are much greater gains to be made with Field-Programmable Gate Array (FPGA) systems.

An FPGA, like a CPU, is a programmable chip that performs computations.  However, they differ in when the program is read.  In a CPU, a program is read and executed as the program runs.  It reads an instruction from an address in memory specified by an instruction counter internal to the CPU.  That instruction will be loaded into the CPU and in executing it, the CPU will change its state, perhaps by loading a value from memory into a CPU register, executing an arithmetic or logic instruction, interfacing with IO, or any of myriad other things.  An FPGA, on the other hand, reads and loads the entire program before execution begins.  It does this by mutating its internal hardware circuits to implement the program.  With this lower-level program implementation an FPGA is able to see higher performance for similar programs.  Instead of indirection through memory lookups, instruction counters, bus loads, and registers, an FPGA can pipeline all necessary instructions in a row.  A PI controller can be implemented as a subtraction circuit, followed by two gains and an accumuluator circuit, followed by an addition circuit.  The circuit would be essentially passive in this case, though there can be memory for more complex logic like would be needed for a PD controller.  The programs for FPGAs are generally written in hardware description languages (HDLs) like Verilog or VHDL which have important hardware concepts like time core to the language, though there are libraries for conventional languages like C++ or Rust to generate HDL code.

The Solid Mechanics Lab employed both a CPU and an FPGA for their control.  The FPGA ran the ``hot'', or high-frequency, calculations, namely the PID loop and also did analog input from the analog-digital converter (ADC).  The ADC actually sampled signals at $\qty{100}{\kilo\hertz}$ and the FPGA program was made to super-sample the input to get an $\qty{18.3}{\bit}$ signal out of a $\qty{16}{\bit}$ ADC.

@@

It also tested this theory against several case studies.  In one case, a hydraulic steering linkage testbed showed that a change in one absraction layer can demand rearchitecture of the other abstraction layers if the control system is not carefully designed.  In another, an organization employed variations on one tech stack to build robots based on soft actuators, presumably to capitalize on institutional knowledge of that stack.  Another used the development of a stewart platform to show how a control system can be developed based on this theory.  The last showed how performance requirements for a strain testing machine can drive the architecture of a control system and how that can be captured by this model.  By organizing the tools used in these case studies based on the function they perform and the interface they provide, I generated these compelling insights into the architecture of the control system.

@@

This theory can also be extended.  Throughout this work, it has been called a \textit{mechatronic} tech stack, but truly it's a control system tech stack.  To make it a true mechatronic tech stack, it should include tooling for mechanical and electronic systems.  A mechanical section of the stack would likely grow from the bottom of the stack shown in figure~\ref{fig:mechatronic_tech_stack}, not colliding with the current stack, but an electrical component would likely be more complicated, since it is needed by both the mechanical stack for transducers and sensors, and by the control stack since the computer \textit{is} an electrical component, and the IO layer often needs some elecrical finesse, as noted in §\ref{sec:stewart_platform}.  It could also be extended in the software realm; the production of graphical monitoring systems and tele-operation programs generally require substantial non-control software which should be described by different layers, perhaps branching away from the control layer to have a parallel monitoring layer.

This theory would also benefit from the study of adjacent work that doesn't develop the mechatronic tech stack itself.  This theory describes the function of mechatronic tooling, but it provides no notion of what tools are best for a particular application.  In evaluating the case studies, some initial guesses were developed, like that tools should encourage loose coupling, take advantage of institutional knowledge, be easy to troubleshoot, and meet certain performance requirments.  This can be formalized by interviewing practitioners and querying what qualities they value in tools as well as by studying extant literature on the subject of tool quality.  These studies would be synthesized into a coherant model for representing the quality of a tool.  Likely this would manifest as a multi-dimensional which a practitioner would interpret to get the best result as apposed to a one-dimensional quality rating that gives practitioners no agency.

As well as considering the quality of a tool, it's important also to study the legal existance of mechatronic tools.  Almost always when making and redistributing mechatronic systems and related design assets, engineers and their associated organizations must agree to some license agreement with the authors of their tools, with these legal agreements being underpinned by various intelectual property laws.  While it is tempting to just ``hit agree'', these license agreements can place notable conditions on the redistribution of mechatronic systems.  A notable subset of these agreements are those \textit{open-source} licenses, which are often discussed as interesting to practitioners in mechatronics, but the legal requirements for open-source software are often ignored or misunderstood.  Open-hardware licenses are even less understood.  A deep dive into law around mechatronic tools would greatly enhance the mechatronic tech stack theory by allowing practitioners to evaluate if they're even allowed to use a tools, regardless of its quality.

>>>>

In order to study mechatronics systems, it is necessary to design and develop physical research test beds. This is especially the case in mechatronic systems with fluid components, as their high-fidelity modeling is challenging and computationally expensive. When doing research with hardware systems, before a single research question can be asked of the hardware, researchers must assemble mechanical components, hook up power electronics, and route hoses for the hydraulic and pneumatic systems. Then the most complex of them all, the control system, must be added. Its tangle of spindly wires reaches all throughout the machine, meeting at the computer, the single most complex component of any fluid power system. The requirements placed on it are immense. In high-power systems, safety invariants must be held up perfectly lest disaster strike, and close monitoring is necessary due to the low stiffness of fluid systems compared to mechanical ones.

@@

Dolores is a full-scale test bed in Aalto University's Fluid Power Laboratory, seen in figure~\ref{fig:dolores0} to study the application of electric hydrostatic actuation in an articulated-steering linkage for a front-loading mining vehicle~\cite{Hermansson2021, Helduser1999, Dudzinski1989}. Its control system commissioning was completed in 2021, but come 2023, an industry partner in the experiment requested hardware changes that demanded a substantial re-architecture of the control system. This case study will discuss the mechatronic tech stacks behind the old and the new systems.

@@

Also interesting in the architecture of a control system is what development environments can be constructed for it.  The development environment is distinct from the heretofore discussed mechatronic stack since it is not a part of the function of the running system, but instead is means which an engineer uses to turn an idea into assets for a mechatronic system.  For a given tech stack, any development environment generally can in principle yield the same assets, but a good tech stack may lend itself to higher quality, more consistent output, faster development cycles, or at least to be enjoyable for individual engineers to use.

For the control layer, I wrote the code in the Sublime Text text editor~\cite{SublimeText} and configured it to automatically format the source code according to standards for the Rust language using the \verb|rustfmt| utility~\cite{rustfmt}.  Other text editors like Emacs, Vim, or VSCodium would have also been good choices, but Sublime is what is easiest for me.  This was done on a Windows workstation, but it was compiled and run on the Linux control computer.  To achieve this, I checked the changes on the workstation into a \verb|git| version control repository and pushed them to the control computer\footnote{A small issue that comes up is that \verb|git| generally doesn't allow users to push changes to a branch of a repository that the local user has checked out, which is the case here~\cite{GitConfigReceiveDenyCurrentBranch}.  This is to prevent data loss.  To enable this, \verb|git| can be configured to update the local branch only if there is no possibility of data loss by executing \verb|git config receive.denyCurrentBranch updateInstead|.}.  Then, though a remote terminal, I executed commands to compile and run the program.  This became quite annoying with how many steps there are, so I use the utility \verb|cargo-watch| to automatically execute the compile and run commands on the control computer when there are updates\footnote{The command was similar to \verb|cargo watch -x 'run -- eth0'|.}~\cite{CargoWatch}.  This still is somewhat tedious and so requires more development.  It also would have been mitigated by developing in some kind of Linux environment or by using a containerized environment like would be provided by Docker or Vagrant, which would have allowed running some tests on the development machine~\cite{DockerDev,VagrantDev}.  Most of the middleware was prepared with the Raspberry Pi OS image that was installed while commissioning the control computer, but there was some additional configuration, e.g.\ to install \verb|git|, that was done with the OS's \verb|apt| package management utility~\cite{apt}.  As well, library dependencies for the control program like Ethercrab were managed with the Rust \verb|cargo| utility~\cite{Cargo}.  The computer, once commissioned, did not require any development.  The IO layer's development was aided by a four-channel oscilloscope and a multimeter to validate electrical connections and ensure signal were propagating as expected.  It was also interesting to observe the timing jitter of the program, which was generally very small, on the order of microseconds.

The in-depth consideration of the development environment was possible because of the open nature of all the tools considered.  Every tool, both a part of the control system and the development environment, have interfaces that enable and encourage users to compose or develop tools to make an environment that meets their needs.  This same discussion could not be had with regards to Dolores since its PLC-based system was very closed off, and any interfacing demanded mastering of low-level communication protocols.  The development environment was simply whatever Beckhoff blessed; no more, no less.

@@

The Solid Mechanics Laboratory at Aalto University conducts experiments to identify the properties of engineering materials under load particularly studying phenomena ranging from the micron to the meter scale.  As is typical for laboratories studying solid mechanics, they do this by subjecting test samples to a load schedule using a servo-controlled hydraulic cylinder.  They have several test beds, each with an instance of a common control system.  This had been controlled by a workstation-commanded DAQ that communicated with a motor controller and a strain gauge amplifier, but new experiments and the desire for faster, better research cycles called for improvements to the control system.

The requirements that drive the design of this control system are mainly related to timing.  There are three compelling design parameters when it comes to timing.  The first is frequency or PID speed; how often in time do control and measurement samples occur?  According to Nyquist, the higher the frequency, the better the control performance~\cite{Nyquist1928}.  Latency, or how old a measurement sample is when it makes it to the PID loop, measured in samples or PID cycles, is also important to control performance.  Thoen shows that in some cases, a system with low sample frequency but low latency can outperform those with high sample frequency and high latency~\cite{Thoen2021}.  Jitter specifies how accurately the sample frequency is kept.  It doesn't impact control performance much, but for high-frequency systems, jitter can be important for compatibility to ensure synchronous operation.  Its specification and control is generally the domain of analog circuit specialists instead of mechatronic generalists.

@@

The requirement for the IO system is that it be low latency.  Up to here the IO latency has been presented as given, but in reality the latency and PID
frequency specifications would be driven by the IO latency.  There are two separate IO systems, one for the strain gauges, which require specialized data acquisition to afford the quality of data needed for solid mechanics experiments.  Each instance of the control system has an HBK QuantumX amplifier, which introduces $\qty{0.3}{\milli\second}$ of latency~\cite{HBKQuantumX}.  For other IO, like signals to and from the servo motor driving the hydraulic pump or monitoring the state of the hydraulic circuit, there needed to be some analog IO ports.  The analog/digital and digital/analog converters contribute notable latency.  For the NI C-Series IO modules that were selected, analog output contributed $\approx\qty{6}{\micro\second}$ of latency per output channel and analog input added $\approx\qty{58}{\micro\second}$ of latency per two channels~\cite{NI-9220,NI-9266}.  This combined in some multiples and the IO, with conditioning, was measured to have the previously stated $ \qtyrange[range-units=single,range-phrase=..]{200}{400}{\micro\second}$ of latency.

The computer system Is interesting to discuss on account of the relatively high PID frequency requirement.  A frequency of $\qty{5}{\kilo\hertz}$ means that all computations for a loop must fit within $\qty{.2}{\milli\second}$.  Each CPU instruction for Intel micro-architectures usually takes ones or tens of clock cycles to execute, or even hundreds for heavy instructions~\cite{Abel19a}.  This means that on a computer with a $\qty{1.3}{\giga\hertz}$ clock like that which this thesis is being written with there can only be some thousands of instructions run per PID cycle.  This will include very fast arithmetic instructions like the additions or multiplications needed for the PID computation, but also commands to fetch data from the CPU's IO ports, which tend to be relatively slow since the signals need to be read and moved into a CPU register or memory location which have their own latency separate from the instruction latency.  On an embedded Linux or Windows system, all this would have to occur while sharing CPU time with other processes.  This isn't to say it's not possible to execute this with embedded Linux or Windows---Beckhoff and NI both provide off-the-shelf solutions for this---but performance issues can often distract from implementing business logic.  This can be mitigated with a bare metal system like was used for the above soft robots, but there are much greater gains to be made with Field-Programmable Gate Array (FPGA) systems.

An FPGA, like a CPU, is a programmable chip that performs computations.  However, they differ in when the program is read.  In a CPU, a program is read and executed as the program runs.  It reads an instruction from an address in memory specified by an instruction counter internal to the CPU.\  That instruction will be loaded into the CPU and in executing it, the CPU will change its state, perhaps by loading a value from memory into a CPU register, executing an arithmetic or logic instruction, interfacing with IO, or any of myriad other things.  An FPGA, on the other hand, reads and loads the entire program before execution begins.  It does this by mutating its internal hardware circuits to implement the program.  With this lower-level program implementation an FPGA is able to see higher performance for similar programs.  Instead of indirection through memory lookups, instruction counters, bus loads, and registers, an FPGA can pipeline all necessary instructions in a row.  A PI controller can be implemented as a subtraction circuit, followed by two gains and an accumulator circuit, followed by an addition circuit.  The circuit would be essentially passive in this case, though there can be memory for more complex logic like would be needed for a PD controller.  The programs for FPGAs are generally written in hardware description languages (HDLs) like Verilog or VHDL which have important hardware concepts like time core to the language, though there are libraries for conventional languages like C++ or Rust to generate HDL code.

The Solid Mechanics Lab employed both a CPU and an FPGA for their control.  The FPGA ran  analog input from the analog-digital converter (ADC) and the ``hot'', or high-frequency, calculations, namely the PID loop.  The ADC actually sampled signals at $\qty{100}{\kilo\hertz}$ and the FPGA program was made to super-sample the input to get an $\qty{18.3}{\bit}$ signal out of a $\qty{16}{\bit}$ ADC.\

At the middleware layer, there is a real-time LabView runtime running atop NI's own real-time Linux distribution~\cite{NICompactRIO}.  The OS includes a module to interface with the FPGA system for data acquisition.  The QuantumX system also brings on its own middleware, catman, which is used for configuration, like to input calibration curves, and for its own data acquisition~\cite{HBKcatman}.  However, in this system, it's merely used as an amplifier.

The control layer was built, like the lower layers, with NI tooling.  All of it was built with LabView, with the CPU-based controls using the LabView Real-Time Module~\cite{NIRealTimeModule} and the FPGA controls built with the LabView FPGA Module~\cite{NIFPGAModule}.  It implemented a PID controller, though the derivative term was in practice usually omitted.  To monitor the condition of long-running tests and for data acquisition, there are programs running on a desktop workstation that observe the state of the control system.  They also developed a custom scripting language that runs on the workstation to define the high-level testing schedule, separate from the low-level control logic.

The Aalto Solid Mechanics Laboratory's new control system

@@

It also tested this theory against several case studies.  In one case, a hydraulic steering linkage test bed showed that a change in one abstraction layer can demand re-architecture of the other abstraction layers if the control system is not carefully designed.  In another, an organization employed variations on one tech stack to build robots based on soft actuators, presumably to capitalize on institutional knowledge of that stack.  Another used the development of a Stewart platform to show how a control system can be developed based on this theory.  The last showed how performance requirements for a strain testing machine can drive the architecture of a control system and how that can be captured by this model.  By organizing the tools used in these case studies based on the function they perform and the interface they provide, I generated these compelling insights into the architecture of the control system.

@@

This theory can also be extended.  Throughout this work, it has been called a \textit{mechatronic} tech stack, but truly it's a control system tech stack.  To make it a true mechatronic tech stack, it should include tooling for mechanical and electronic systems.  A mechanical section of the stack would likely grow from the bottom of the stack shown in figure~\ref{fig:mechatronic_tech_stack}, not colliding with the current stack, but an electrical component would likely be more complicated, since it is needed by both the mechanical stack for transducers and sensors, and by the control stack since the computer \textit{is} an electrical component, and the IO layer often needs some electrical finesse, as noted in §\ref{sec:stewart_platform}.  It could also be extended in the software realm; the production of graphical monitoring systems and tele-operation programs generally require substantial non-control software which should be described by different layers, perhaps branching away from the control layer to have a parallel monitoring layer.

This theory would also benefit from the study of adjacent work that doesn't develop the mechatronic tech stack itself.  This theory describes the function of mechatronic tooling, but it provides no notion of what tools are best for a particular application.  In evaluating the case studies, some initial guesses were developed, like that tools should encourage loose coupling, take advantage of institutional knowledge, be easy to troubleshoot, and meet certain performance requirements.  This can be formalized by interviewing practitioners and querying what qualities they value in tools as well as by studying extant literature on the subject of tool quality.  These studies would be synthesized into a coherent model for representing the quality of a tool.  Likely this would manifest as a multi-dimensional which a practitioner would interpret to get the best result as apposed to a one-dimensional quality rating that gives practitioners no agency.

As well as considering the quality of a tool, it's important also to study the legal existence of mechatronic tools.  Almost always when making and redistributing mechatronic systems and related design assets, engineers and their associated organizations must agree to some license agreement with the authors of their tools, with these legal agreements being underpinned by various intellectual property laws.  While it is tempting to just ``hit agree'', these license agreements can place notable conditions on the redistribution of mechatronic systems.  A notable subset of these agreements are those \textit{open-source} licenses, which are often discussed as interesting to practitioners in mechatronics, but the legal requirements for open-source software are often ignored or misunderstood.  Open-hardware licenses are even less understood.  A deep dive into law around mechatronic tools would greatly enhance the mechatronic tech stack theory by allowing practitioners to evaluate if they're even allowed to use a tools, regardless of its quality.
